Scan-Centric, Frequency-Based Method for Characterizing Peaks from Direct Injection Fourier transform Mass Spectrometry Experiments

### 2022-06-01 doc/peakcharacterization_manuscript.Rmd
some minor edits

::: {custom-style="MDPI_1.7_abstract"}
**Abstract:** We present a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw m/z points in spectral scans.
Our peak characterization method utilizes intensity independent noise removal and normalization of scan-level data to provide a much better fit of relative intensity to natural abundance probabilities for low abundance isotopologues that are not present in all of the acquired scans.
Moreover, our method calculates both peak- and scan-specific statistics incorporated within a series of quality control steps that are designed to robustly derive peak centers, intensities and intensity ratios with their scan-level variances.
These cross-scan characterized peaks are suitable for use in our previously published peak assignment methodology, Small Molecule Isotope Resolved Formula Enumeration (SMIRFE).
:::

::: {custom-style="MDPI_1.8_keywords"}
**Keywords**: Fourier-transform mass spectrometry, orbitrap, scan-centric peak characterization, Fellgett's advantage, interferograms, frequency spectrum
:::

@@

::: {custom-style="MDPI_3.1_text"}
Fourier-transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution, and mass accuracy all in one analytical instrumentation.
The combination of these capabilities provide several analytical and interpretive improvements: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enabling multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007]. 
In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of single and multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017].
These stable isotope tracing experiments provide a wealth of isotope flux data that is interpretable in terms of metabolite flux information that is specific to a metabolic model, pathway, and subcellular location [@higashiStableIsotopeLabeledTracers2014; @creekStableIsotopeAssistedMetabolomics2012; @hillerNontargetedElucidationMetabolic2010; @fanStableIsotoperesolvedMetabolomics2012; @moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017; @jinMoietyModelingFramework2019; @jinRobustMoietyModel2020].

@@

The input peak data in these studies are derived from multi-scan direct-injection FT-MS data.
SMIRFE uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peak pairs to determine the likelihood of an assignment for the peaks in question.
In the Thermo-Fisher Orbitrap FT-MS instruments, data is acquired in microscans and scans, where each scan is an aggregate of multiple microscans.
While these microscan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e., "bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak m/z and intensity.
If the scan level data is not processed and aggregated correctly, the final peak intensities in arbitrary units and centers in m/z will have high uncertainty (total variance) [@moseleyErrorAnalysisPropagation2013].

In particular, if peaks are missing in some scans due to low abundance, the final aggregate peak intensities will not fit the expected relative intensity to NAP relationships.
Here we describe novel scan-centric FT-MS metabolomics data processing methods that better preserve the expected relative intensity to NAP.
In addition, due to the tendency of increasing point spacing with increasing m/z in Orbitrap type instruments, we derive a method for transforming the m/z data to axial frequency, which has the desirable property of being equally spaced across the full spectrum.
Finally, we show that in addition to better preserving the expected intensity to NAP relationships, our scan-centric method also results in improved relative standard deviations, automatic identification of high-peak-density artifact peaks, and better separation of samples in a lipidomics data analysis.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("motivation")`.**
**A**: Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
The 18O isotopologue is shifted by 0.01 m/z and the NAP peaks by 0.05 m/z to aid visualization.
**B**: Theoretical NAP-based-Log10-intensities and Log10 Xcalibur intensities.
The red line indicates perfect agreement.
:::

::: {custom-style="MDPI_2.2_heading2"}
### 2.2. m/z to Frequency
:::

@@

::: {custom-style="MDPI_3.1_text"}
FT-MS data from the Thermo-Fisher Orbitrap instruments used to acquire the data does not provide any information about the raw spectral frequency data.
Outside of the meta-data, it merely contains the m/z and intensity values for profile spectra acquired across multiple scans.
However, the spectral frequency can be calculated by dividing the midpoint m/z of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the midpoint m/z of two adjacent points, length of red lines representing the difference between the two adjacent points).
The subsequent adjacent point differences in frequency are expected to be relatively constant with respect to m/z, as shown in `r figure_count$label_text("mz_frequency_conversion")`C, `r figure_count$just_count("mz_frequency_conversion")`D and `r figure_count$just_count("mz_frequency_conversion")`E, in contrast to the adjacent point differences in m/z, which are not constant with respect to m/z.
The Thermo-Fisher Fusion instrument from which most of our collaborators' data has been acquired, at a resolution of 450K or 500K depending on the sample, has an adjacent frequency point difference mode of 0.5, as shown in  `r figure_count$label_text("mz_frequency_conversion")`D.
It is important to note that although the frequency difference has been consistent regardless of instrument resolution, we do not know a priori what it should be, and the method makes no assumptions about the frequency difference, but rather calculates the mode of the frequency differences from those observed in the data itself.
Restricting to those points that fall into a narrow range around the mode of frequency differences (0.49 - 0.51 in this work), a regression model of frequency to m/z can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`F.
This regression model seems to fit the known relationship between frequency and m/z, where the frequency is related to $1/\sqrt{m/z}$.
The actual regression model used in this work includes terms for m/z, $\sqrt{m/z}$, and $\sqrt[3]{m/z}$ (see **Conversion of m/z to Frequency** in Methods for further description of the regression model).
We did investigate a variety of frequency regression models to see which one seemed to be the best, as shown in Supplemental `r supp_figure_count$label_text("frequency_models")` and `r supp_table_count$label_text("frequency_models")`.

The constant point difference of frequency space is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
The m/z point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is exceedingly difficult to create a LOESS model with an intercept of 0.
In addition, we would also need to vary the width of sliding windows according to the m/z difference at a particular m/z based on the LOESS model.
Frequency-based points suffer none of these drawbacks, and the conversion from m/z can be derived from the raw profile level data itself, which is incredibly useful.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("mz_frequency_conversion")`.** 
**A**: Intensity vs m/z for a single peak from a single scan.
Red lines denote the difference between each adjacent pair of points, and red dots the m/z midpoint between the pair of points.
The difference divided by the midpoint is used to derive the spectral frequency values in **B**.
**B** plots the intensity vs the converted frequency points derived from **A**.
The red lines denote the difference between adjacent points, which are shown in the y-axis of **C** for this single peak.
**D** shows a histogram of adjacent point differences for all points. Note that the y-axis is Log10 scaled.
The differences for all points in a single scan vs m/z are shown in **E**, with those differences that lie within 0.49 - 0.51 shown in red.
**F** shows the plot of derived spectral frequency vs m/z, with fitted values from the linear regression model in red.
:::

@@

::: {custom-style="MDPI_3.1_text"}
The m/z to frequency regression models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
While scan specific models **could** be used in the conversion of m/z to frequency, doing so results in changes to the relative peak ordering compared to m/z space, as shown in `r figure_count$label_text("peak_ordering")`.
Therefore, a single model for all scans based on the scan with the slope closest to the median of slopes across scans is used for converting *all* remaining scan level data.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("peak_ordering")`.** Peak ordering in m/z compared with ordering in frequency space when a single m/z to frequency model is used or scan specific m/z to frequency models are used.
For a single peak, the scan level peak m/z's were extracted, and then frequency values for those m/z generated using a single common model of m/z to frequency (*Single Model*), or models derived from each scan (*Individual Models*).
A subset of the peaks ends up out of order using scan specific models, implying that a single model should be used across all the scan level data.
:::

::: {custom-style="MDPI_3.1_text"}
Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** m/z points are converted to frequency for subsequent steps in the workflow.
:::

@@

::: {custom-style="MDPI_3.1_text"}
In a dataset of this nature, we expect that much of the data is really just noise, and does not contribute meaningfully to the analysis.
Furthermore, it is expected that noise is randomly distributed across the scans.
Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
Subsequently, we divide the counts into tiled regions of a set size (1000 frequency in this work), and examine the cumulative 99^th^ percentile of the sliding window counts within the tiled region.
In `r figure_count$label_text("slidingwindow_count")`A, we plotted the non-zero point intensities from all scans for one tiled region 1000 frequency wide.
The number of non-zero points in the sliding windows across this region are shown in `r figure_count$label_text("slidingwindow_count")`B, and zoomed closer in `r figure_count$just_count("slidingwindow_count")`C.
These provide a measure of the non-zero point density across this region.
The blue line indicates the cumulative 99^th^ percentile for this one region, and is high, given that it looks like there are some consistent peaks across the scans in this region.
However, examining the cumulative 99^th^ percentiles across all the regions of the spectrum, as shown in the histogram of `r figure_count$label_text("slidingwindow_count")`F, a rather large fraction of the regions have very low cumulative 99^th^ percentile values, with a median value of 5, which results in a cutoff value of 8 (ceiling of 5 x 1.5).
Keeping only those sliding regions with a non-zero point count greater than 8 (above the red line in `r figure_count$label_text("slidingwindow_count")`B and `r figure_count$just_count("slidingwindow_count")`C) results in those sliding regions shown in `r figure_count$just_count("slidingwindow_count")`D, and then the point intensities in `r figure_count$just_count("slidingwindow_count")`E.
A non-zero point density based method for removing potential noise is desirable because the peak intensities across scans are not identical, as shown by the need for normalization (see **Normalization of Scans**), and the distribution of RSD values before and after normalization (see **Changes in Relative Standard Deviation**).
:::

@@

::: {custom-style="MDPI_3.1_text"}
Although many other types of mass spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion has a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier. 
For each region initially created, the peaks in each scan within that region can be characterized (i.e., centroiding).
For centroiding, we use a simple weighted quadratic model of log(intensity) to m/z. 
We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of peak, helping to ensure that the **true** centroid is determined. 
From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in `r figure_count$label_text("centroided_peaks")`.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("centroided_peaks")`.**
Log-intensity (A) and Intensity (B) of points for a single peak against frequency. Black points are the original data points, and the blue point represents the calculated centroid.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("breaking_regions")`.** Splitting a single region into two regions based on the peaks that are present.
**A**: The full set of raw frequency and intensity data across all scans for the region are shown.
Each horizontal trace are the point intensities in a single scan.
Clearly the region has two separate peaks within it.
**B**: The peak centroids (frequency and intensity) for each peak in black.
The tiled regions (red) are used to quantify the number of peaks.
**C**: The number of peaks within each tiled region shown as a histogram.
Each group of non-zero adjacent regions will be merged to form a new peak region.
:::

@@

::: {custom-style="MDPI_3.1_text"}
Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary between scans (see `r figure_count$label_text("breaking_regions")`A and `r figure_count$just_count("breaking_regions")`B for example).
Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
However, there are two issues with these peak intensities across scans: (a) some peak intensities correlate with the scan number (i.e., scan acquisition order); and (b) some peak differences between scans are correlated with intensity.
The solution to **a** is to do a two-pass normalization.
After the first pass, the peaks whose intensity is correlated with scan order are detected (absolute Pearson correlation > 0.5).
In the second pass, the correlated peaks are removed, and normalization is carried out again.
`r figure_count$label_text("intensity_scan")`A shows an example peak whose intensity across scans is correlated with scan number.
The solution to **b** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and `r figure_count$just_count("normalization_factors")`B.
Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B), which is expected as they both use the most intense peaks. 
But removing the scan-correlated peaks does change the majority of the normalization factors.
:::

::: {custom-style="MDPI_5.2_figure"}
```{r intensity_scan, dn_id = figure_count}
# showing peak intensity correlation with m/z, intensity varying with m/z
tar_load(intensity_scan_cor_plot)
intensity_scan_cor_plot
```
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single-pass normalization using *all* peaks (singlenorm), single-pass normalization using peaks with an intensity ≥ 0.7 of the maximum intensity (singlenorm_int), or the two-pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (doublenorm).
**B**: The difference of the normalization factors obtained from either doublenorm or singlenorm_int compared to singlenorm.
:::

@@

::: {custom-style="MDPI_3.1_text"}
We have previously described the presence of high peak density (HPD) artifacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
Ideally, the peak characterization procedure should reduce and mitigate their presence and effect in the resultant reported peaks.
Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan.
However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
To verify the mitigation of HPD regions, we converted centroided m/z's from Xcalibur to frequency values using the previously calculated regression model for that sample, measured peak density to detect HPD regions, and compared them with the scan-centric peaks and FSD outliers.
The Xcalibur peak density was measured using a sliding window ten points wide and a stride of one point.

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("rsd_method")`.**
Density plots of relative standard deviations (RSD) of peak heights across scans for each of the processing methods.
A peak had to be present in at least three scans for the RSD value to be reported.
:::

@@

::: {custom-style="MDPI_3.1_text"}
The 97lipid filtersd RSD distribution (as well as others) is bi-modal.
`r figure_count$label_text("rsd_intensity")` plots each peaks RSD as a function of the mean intensity across scans, for every peak present in at least three scans (top), or the peaks present in at least 80% of the scans (bottom).
At least part of the bi-modality in the RSD distribution appears to be related to this dependence of RSD on intensity, as well as truncation as peaks appear in fewer scans.
This is supported by part of the distribution disappearing when we require that peaks be in at least 80% of the available scans (difference between `r figure_count$label_text("rsd_intensity")` top and bottom).

We can remove a majority of the bi-modality by only examining those peaks with a Log10(mean) intensity ≥ 5.
`r supp_figure_count$label_text("rsd_method")` and `r supp_table_count$label_text("rsd_method")` show how the RSD distributions change for each sample when only the peaks with Log10(mean) intensity ≥ 5 are used.
The distributions are all shifted to lower RSD, however, the overall trends in RSD are the same.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("assigned_mz_diff")`.**
Histograms of the differences between observed and expected m/z values for assigned peaks in the 1ecf (**left**) and 97lipid (**right**) spectra, measured in parts-per-million (ppm) respectively.
Inset plots are scaled to the maximum counts observed for the particular method.
:::

@@

long_pvalue = long_ttest %>%
  dplyr::filter(measure %in% "p.adjust")

>>>>

::: {custom-style="MDPI_1.7_abstract"}
**Abstract:** We present a novel, scan-centric method for characterizing peaks from direct injection multi-scan Fourier transform mass spectra of complex samples that utilizes frequency values derived directly from the spacing of raw *m/z* points in spectral scans.
Our peak characterization method utilizes intensity independent noise removal and normalization of scan-level data to provide a much better fit of relative intensity to natural abundance probabilities for low abundance isotopologues that are not present in all of the acquired scans.
Moreover, our method calculates both peak- and scan-specific statistics incorporated within a series of quality control steps that are designed to robustly derive peak centers, intensities and intensity ratios with their scan-level variances.
These cross-scan characterized peaks are suitable for use in our previously published peak assignment methodology, Small Molecule Isotope Resolved Formula Enumeration (SMIRFE).
:::

::: {custom-style="MDPI_1.8_keywords"}
**Keywords**: Fourier-transform mass spectrometry; orbitrap; scan-centric peak characterization; Fellgett's advantage; interferograms; frequency spectrum
:::

@@

::: {custom-style="MDPI_3.1_text"}
Fourier-transform mass spectrometry (FT-MS) provides high performance in terms of sensitivity, resolution, and mass accuracy all in one analytical instrumentation.
The combination of these capabilities provides several analytical and interpretive improvements: (i) the ability to resolve distinct isotopologues with identical unit masses but different accurate masses [@higashiStableIsotopeLabeledTracers2014]; (ii)  enabling multi-element isotopic natural abundance correction for at least the lower portion of the detected mass range  [@moseleyCorrectingEffectsNatural2010; @carreerComputationalFrameworkHighThroughput2013; @wangAccuCor2IsotopeNatural2021]; (iii) improved assignment accuracy [@kindMetabolomicDatabaseAnnotations2006; @mitchellSmallMoleculeIsotope2019]; and (iv) the detection of metabolites in the sub-femtomolar range, when combined with chromatographic separation  [@eylesMethodsStudyProtein2004; @dettmerMassSpectrometrybasedMetabolomics2007]. 
In the metabolomics field, these improvements permit more complicated, but more informative experimental designs such as the use of single and multiple isotope-labeled precursors in stable isotope-resolved metabolomics (SIRM) experiments [@yangChloroformateDerivatizationTracing2017].
These stable isotope tracing experiments provide a wealth of isotope flux data that is interpretable in terms of metabolite flux information that is specific to a metabolic model, pathway, and subcellular location [@higashiStableIsotopeLabeledTracers2014; @creekStableIsotopeAssistedMetabolomics2012; @hillerNontargetedElucidationMetabolic2010; @fanStableIsotoperesolvedMetabolomics2012; @moseleyNovelDeconvolutionMethod2011; @sellersPyruvateCarboxylaseCritical2015; @verdegemMAIMSSoftwareTool2017; @jinMoietyModelingFramework2019; @jinRobustMoietyModel2020].

@@

The input peak data in these studies are derived from multi-scan direct-injection FT-MS data.
SMIRFE uses expected relationships of relative peak height or intensity to natural abundance probability (NAP) across multiple peak pairs to determine the likelihood of an assignment for the peaks in question.
In the Thermo-Fisher Orbitrap FT-MS instruments, data are acquired in microscans and scans, where each scan is an aggregate of multiple microscans.
While these microscan and subsequent scans are expected to be analytical replicates, spray instability and temporal delay in automatic gain control pragmatically break this expectation.
Therefore, it is advantageous to keep the data at the scan level, so that any given poor quality (i.e., "bad") scan can be removed prior to aggregating the data across scans and generating a centroided peak *m/z* and intensity.
If the scan level data are not processed and aggregated correctly, the final peak intensities in arbitrary units and centers in *m/z* will have high uncertainty (total variance) [@moseleyErrorAnalysisPropagation2013].

In particular, if peaks are missing in some scans due to low abundance, the final aggregate peak intensities will not fit the expected relative intensity to NAP relationships.
Here we describe novel scan-centric FT-MS metabolomics data processing methods that better preserve the expected relative intensity to NAP.
In addition, due to the tendency of increasing point spacing with increasing *m/z* in Orbitrap type instruments, we derive a method for transforming the *m/z* data to axial frequency, which has the desirable property of being equally spaced across the full spectrum.
Finally, we show that in addition to better preserving the expected intensity to NAP relationships, our scan-centric method also results in improved relative standard deviations, automatic identification of high-peak-density artifact peaks, and better separation of samples in a lipidomics data analysis.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("motivation")`.**
**A**: Xcalibur intensities (black) and theoretical intensities based on relative NAP (red) for ECF derived threonine.
The 18O isotopologue is shifted by 0.01 *m/z* and the NAP peaks by 0.05 *m/z* to aid visualization.
**B**: Theoretical NAP-based-Log10-intensities and Log10 Xcalibur intensities.
The red line indicates perfect agreement.
:::

::: {custom-style="MDPI_2.2_heading2"}
### 2.2. *m/z* to Frequency
:::

@@

::: {custom-style="MDPI_3.1_text"}
FT-MS data from the Thermo-Fisher Orbitrap instruments used to acquire the data does not provide any information about the raw spectral frequency data.
Outside of the meta-data, it merely contains the *m/z* and intensity values for profile spectra acquired across multiple scans.
However, the spectral frequency can be calculated by dividing the midpoint *m/z* of two adjacent points by their difference (`r figure_count$label_text("mz_frequency_conversion")`A, red points representing the midpoint *m/z* of two adjacent points, length of red lines representing the difference between the two adjacent points).
The subsequent adjacent point differences in frequency are expected to be relatively constant with respect to *m/z*, as shown in `r figure_count$label_text("mz_frequency_conversion")`C, `r figure_count$just_count("mz_frequency_conversion")`D and `r figure_count$just_count("mz_frequency_conversion")`E, in contrast to the adjacent point differences in *m/z*, which are not constant with respect to *m/z*.
The Thermo-Fisher Fusion instrument from which most of our collaborators' data has been acquired, at a resolution of 450 K or 500 K depending on the sample, has an adjacent frequency point difference mode of 0.5, as shown in  `r figure_count$label_text("mz_frequency_conversion")`D.
It is important to note that although the frequency difference has been consistent regardless of instrument resolution, we do not know a priori what it should be, and the method makes no assumptions about the frequency difference, but rather calculates the mode of the frequency differences from those observed in the data itself.
Restricting to those points that fall into a narrow range around the mode of frequency differences (0.49 - 0.51 in this work), a regression model of frequency to *m/z* can be generated (see Methods), with an example shown in `r figure_count$label_text("mz_frequency_conversion")`F.
This regression model seems to fit the known relationship between frequency and *m/z*, where the frequency is related to $1/\sqrt{m/z}$.
The actual regression model used in this work includes terms for m/z, $\sqrt{m/z}$, and $\sqrt[3]{m/z}$ (see **Conversion of m/z to Frequency** in Methods for further description of the regression model).
We did investigate a variety of frequency regression models to see which one seemed to be the best, as shown in Supplemental `r supp_figure_count$label_text("frequency_models")` and `r supp_table_count$label_text("frequency_models")`.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("mz_frequency_conversion")`.** 
**A**: Intensity vs *m/z* for a single peak from a single scan.
Red lines denote the difference between each adjacent pair of points, and red dots the *m/z* midpoint between the pair of points.
The difference divided by the midpoint is used to derive the spectral frequency values in **B**.
**B** plots the intensity vs the converted frequency points derived from **A**.
The red lines denote the difference between adjacent points, which are shown in the y-axis of **C** for this single peak.
**D** shows a histogram of adjacent point differences for all points. Note that the y-axis is Log10 scaled.
The differences for all points in a single scan vs *m/z* are shown in **E**, with those differences that lie within 0.49 - 0.51 shown in red.
**F** shows the plot of derived spectral frequency vs *m/z*, with fitted values from the linear regression model in red.
:::

::: {custom-style="MDPI_3.1_text"}
The constant point difference in frequency space is useful, because some of the subsequent steps in our workflow use sliding and tiled windows where it is assumed that the sliding windows contain the same number of data points.
The *m/z* point-to-point differences are not constant, but can be approximated by a LOESS linear model [@Cleveland1992loess]; however, it is exceedingly difficult to create a LOESS model with an intercept of 0.
In addition, we would also need to vary the width of sliding windows according to the *m/z* difference at a particular *m/z* based on the LOESS model.
Frequency-based points suffer from none of these drawbacks, and the conversion from *m/z* can be derived from the raw profile level data itself, which is incredibly useful.
:::

@@

::: {custom-style="MDPI_3.1_text"}
The *m/z* to frequency regression models are calculated for each scan, and the square root term from all scan level models are checked for outliers based on the interquartile ranges across all scans in a sample.
While scan specific models **could** be used in the conversion of *m/z* to frequency, doing so results in changes to the relative peak ordering compared to *m/z* space, as shown in `r figure_count$label_text("peak_ordering")`.
Therefore, a single model for all scans based on the scan with the slope closest to the median of slopes across scans is used for converting *all* remaining scan level data.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("peak_ordering")`.** Peak ordering in *m/z* compared with ordering in frequency space when a single *m/z* to frequency model is used or scan specific *m/z* to frequency models are used.
For a single peak, the scan level peak *m/z*'s were extracted, and then frequency values for those *m/z* generated using a single common model of *m/z* to frequency (*Single Model*), or models derived from each scan (*Individual Models*).
A subset of the peaks ends up out of order using scan specific models, implying that a single model should be used across all the scan level data.
:::

::: {custom-style="MDPI_3.1_text"}
Although the original model is created from only those points that had frequency point-to-point differences within a narrow range, **all** *m/z* points are converted to frequency for subsequent steps in the workflow.
:::

@@

::: {custom-style="MDPI_3.1_text"}
In a dataset of this nature, we expect that much of the data are really just noise, and do not contribute meaningfully to the analysis.
Furthermore, it is expected that noise is randomly distributed across the scans.
Therefore, if we slide a window across the data and sum the number of non-zero points in each window, we expect that most of the data we encounter is actually noise.
Subsequently, we divide the counts into tiled regions of a set size (1000 frequency in this work), and examine the cumulative 99^th^ percentile of the sliding window counts within the tiled region.
In `r figure_count$label_text("slidingwindow_count")`A, we plotted the non-zero point intensities from all scans for one tiled region 1000 frequency wide.
The number of non-zero points in the sliding windows across this region are shown in `r figure_count$label_text("slidingwindow_count")`B, and zoomed closer in `r figure_count$just_count("slidingwindow_count")`C.
These provide a measure of the non-zero point density across this region.
The blue line indicates the cumulative 99^th^ percentile for this one region, and is high, given that it looks like there are some consistent peaks across the scans in this region.
However, examining the cumulative 99^th^ percentiles across all the regions of the spectrum, as shown in the histogram of `r figure_count$label_text("slidingwindow_count")`F, a rather large fraction of the regions have very low cumulative 99^th^ percentile values, with a median value of 5, which results in a cutoff value of 8 (ceiling of 5 x 1.5).
Keeping only those sliding regions with a non-zero point count greater than 8 (above the red line in `r figure_count$label_text("slidingwindow_count")`B and `r figure_count$just_count("slidingwindow_count")`C) results in those sliding regions shown in `r figure_count$just_count("slidingwindow_count")`D, and then the point intensities in `r figure_count$just_count("slidingwindow_count")`E.
A non-zero point density based method for removing potential noise is desirable because the peak intensities across scans are not identical, as shown by the need for normalization (see **Normalization of Scans**), and the distribution of RSD values before and after normalization (see **Changes in Relative Standard Deviation**).
:::

@@

::: {custom-style="MDPI_3.1_text"}
Although many other types of mass spectrometry data suffer from a variable and noisy baseline, the scan-level profile data from the Thermo-Fisher Fusion have a baseline of 0 due to manipulations in the Thermo-Fisher firmware, making the determination of the centroided values considerably easier. 
For each region initially created, the peaks in each scan within that region can be characterized (i.e., centroiding).
For centroiding, we use a simple weighted quadratic model of log(intensity) to *m/z*. 
We add a small constant to enable using the zero values, and weight the values by their ratio to the most intense value, which is normally the value closest to the center of the peak, helping to ensure that the **true** centroid is determined. 
From the fitted model, we can derive the centroided center and the intensity of the peak, as shown in `r figure_count$label_text("centroided_peaks")`.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("centroided_peaks")`.**
Log-intensity (A) and intensity (B) of points for a single peak against frequency. Black points are the original data points, and the blue point represents the calculated centroid.
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("breaking_regions")`.** Splitting a single region into two regions based on the peaks that are present.
**A**: The full set of raw frequency and intensity data across all scans for the region are shown.
Each horizontal trace are the point intensities in a single scan.
Clearly the region has two separate peaks within it.
**B**: The peak centroids (frequency and intensity) for each peak are in black.
The tiled regions (red) are used to quantify the number of peaks.
**C**: The number of peaks within each tiled region are shown as a histogram.
Each group of non-zero adjacent regions will be merged to form a new peak region.
:::

@@

::: {custom-style="MDPI_3.1_text"}
Due to differences in how many ions are captured in the trap and the limited dynamic range of the detector, the observed peak intensities for the same analyte vary between scans (see `r figure_count$label_text("breaking_regions")`A and `r figure_count$just_count("breaking_regions")`B for example).
Using the median peak differences between scans, it is possible to normalize the peak intensities across scans.
However, there are two issues with these peak intensities across scans: (a) some peak intensities correlate with the scan number (i.e., scan acquisition order); and (b) some peak differences between scans are correlated with intensity.
The solution to **a** is to do a two-pass normalization.
After the first pass, the peaks whose intensity is correlated with scan order are detected (absolute Pearson correlation > 0.5).
In the second pass, the correlated peaks are removed, and normalization is carried out again.
`r figure_count$label_text("intensity_scan")`A shows an example peak whose intensity across scans is correlated with scan number.
The solution to **b** (peak differences correlated with intensity) is to only use the most intense peaks, as shown in `r figure_count$label_text("intensity_scan")`B.
The highlighted peaks in `r figure_count$label_text("intensity_scan")`B are those with an intensity greater than 0.7 of the maximum intensity observed in that scan, and at least visually, their differences are **not** correlated with intensity.
If *all* peaks are used for normalization, a very different set of normalization factors will be generated than by using only the *most intense* peaks, as shown in `r figure_count$label_text("normalization_factors")`A and `r figure_count$just_count("normalization_factors")`B.
Though the two-pass normalization (doublenorm) and single-pass normalization using the most intense peaks (singlenorm_int) generate very similar normalization factors (`r figure_count$label_text("normalization_factors")`B), this is expected as they both use the most intense peaks. 
However, removing the scan-correlated peaks does change the majority of the normalization factors.
:::

::: {custom-style="MDPI_5.2_figure"}
```{r intensity_scan, dn_id = figure_count}
# showing peak intensity correlation with *m/z*, intensity varying with *m/z*
tar_load(intensity_scan_cor_plot)
intensity_scan_cor_plot
```
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("normalization_factors")`.** **A**: Histogram of scan normalization factors using either a single-pass normalization using *all* peaks (*singlenorm*), single-pass normalization using peaks with an intensity ≥ 0.7 of the maximum intensity (*singlenorm_int*), or the two-pass normalization removing peaks whose height is correlated with scan and using the most intense peaks (*doublenorm*).
**B**: The difference of the normalization factors obtained from either *doublenorm* or *singlenorm_int* compared to *singlenorm*.
:::

@@

::: {custom-style="MDPI_3.1_text"}
We have previously described the presence of high peak density (HPD) artifacts in FT-MS spectra [@mitchellNewMethodsIdentify2018].
Ideally, the peak characterization procedure should reduce and mitigate their presence and effect in the resultant reported peaks.
Their presence should be minimized by removing noise peaks, and removing peak regions that have multiple reported peaks from the same scan.
However, we expect they may also present as characterized peaks that have larger than expected frequency level standard deviations (FSD) when calculated across scans.
These peaks can be detected by simply examining the distribution of FSDs and marking outlier peaks.
To verify the mitigation of HPD regions, we converted centroided *m/z*'s from Xcalibur to frequency values using the previously calculated regression model for that sample, measured peak density to detect HPD regions, and compared them with the scan-centric peaks and FSD outliers.
The Xcalibur peak density was measured using a sliding window ten points wide and a stride of one point.

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("rsd_method")`.**
Density plots for relative standard deviations (RSD) of peak heights across scans for each of the processing methods.
A peak was required to be present in at least three scans for the RSD value to be reported.
:::

@@

::: {custom-style="MDPI_3.1_text"}
The 97lipid filtersd RSD distribution (as well as others) is bi-modal.
`r figure_count$label_text("rsd_intensity")` plots each peak RSD as a function of the mean intensity across scans, for every peak present in at least three scans (top), or the peaks present in at least 80% of the scans (bottom).
At least part of the bi-modality in the RSD distribution appears to be related to this dependence of RSD on intensity, as well as truncation as peaks appear in fewer scans.
This is supported by part of the distribution disappearing when we require that peaks be in at least 80% of the available scans (difference between `r figure_count$label_text("rsd_intensity")` top and bottom).
:::

@@

::: {custom-style="MDPI_5.1_figure_caption"}
**`r figure_count$label_text("assigned_mz_diff")`.**
Histograms of the differences between observed and expected *m/z* values for assigned peaks in the 1ecf (**left**) and 97lipid (**right**) spectra, measured in parts-per-million (ppm) respectively.
Inset plots are scaled to the maximum counts observed for the particular method.
:::

@@

scientific_10 = function(x) {
  gsub("e\\+*", " %*% 10^", scales::scientific_format()(x))
}
